\bibliographystyle{plain}
\begin{thebibliography}{99}

%% \bibitem{SNMP}
%% Case, J.~D., Fedor, M., and Schoffstall, M.~L.
%% ``Simple Network Management Protocol (SNMP).'' Internet Request for
%% Comments~1157, May 1990.

\bibitem{ComerV1}
Comer, Douglas E. \emph{Internetworking with TCP/IP,} volume~1,
\emph{Principles, Protocols, and Architecture} (third edition).
Prentice Hall, 1995.

\bibitem{ComerV2}
Comer, Douglas E., and Stevens, David L. \emph{Internetworking with
TCP/IP,} volume~2, \emph{Design, Implementation, and Internals} (third
edition).  Prentice Hall, 1999.

\bibitem{ComerV3}
Comer, Douglas E., and Stevens, David L. \emph{Internetworking with
TCP/IP,} volume~3, \emph{Client-Server Programming and Applications}
(BSD version, second edition).
Prentice Hall, 1996.

\bibitem{RFC4291}
Hinden, R., and Deering, S.
``IP Version 6 Addressing Architecture'', Internet Request for
Comments~4291, February 2006.

\bibitem{RFC2460}
Deering, S., and Hinden, R. ``Internet Protocol, Version 6 (IPv6)
Specification.'' Internet Request for Comments 2460, December 1998.

\bibitem{RFC3493}
Gilligan, R., Thomson, S., Bound, J., McCann, J., and Stevens, W.
``Basic Socket Interface Extensions for IPv6.''
Internet  Request for Comments 3493, February 2003.

\bibitem{RFC3022}
Srisuresh, P., Egevang, S.
``Traditional IP Network Address Translator (Traditional NAT).''
Internet  Request for Comments 3022, January 2001.

%% \bibitem{ASN1}
%% International Organization for Standardization. \emph{Basic Encoding
%% Information Processing Systems: Open Systems Interconnection---Specification
%% of Abstract Syntax Notation One (ASN.1)}.
%% International Standard 8824, December 1987.

\bibitem{RFC1034}
Mockapetris, Paul.
``Domain Names: Concepts and Facilities.''
Internet  Request for Comments 1034, November 1987.

\bibitem{RFC1035}
Mockapetris, Paul.
``Domain Names: Implementation and Specification.''
Internet  Request for Comments 1035, November 1987.

\bibitem{PandD}
Peterson, Larry L., and Davie, Bruce S.
\emph{Computer Networks: A Systems Approach} (fourth edition).
Morgan Kaufmann, 2007.

\bibitem{RFC4038}
 M-K. Shin, et al.
``Application Aspects of IPv6 Transition'', Internet Request for
Comments 4038, March 2005

\bibitem{RFC791}
Postel, John. ``Internet Protocol.'' Internet Request for Comments
791, September 1981.

\bibitem{RFC793}
Postel, John. ``Transmission Control Protocol.'' Internet Request for Comments
793, September 1981.

\bibitem{RFC768}
Postel, John. ``User Datagram Protocol.'' Internet Request for Comments
768, August 1980.

\bibitem{StevensV1}
Stevens, W. Richard. \emph{TCP/IP Illustrated,} volume~1, \emph{The Protocols}.
Addison-Wesley, 1994.

\bibitem{StevensUNP}
Stevens, W. Richard. \emph{UNIX Network Programming: Networking
APIs: Sockets and XTI} (second edition). Prentice Hall, 1997.

\bibitem{StevensV2}
Wright, Gary  R., and Stevens, W. Richard. \emph{TCP/IP Illustrated,}
volume~2, \emph{The Implementation}. Addison-Wesley, 1995.

\end{thebibliography}
